Image Caption Generation With CLIP+GPT-2 Model

### 2025-05-04 Latex_Paper/Paper.tex
update all

Motivated by these developments, and after reviewing related 
literature \cite{Nukrai2022},we aim to replicate a CLIP+GPT-2 based 
image captioning system. Through this project, we seek to explore and deepen our understanding of 
machine learning, as well as its broader applications in fields such as computer vision and natural language generation.
\begin{itemize}
    \item Leverage the pretrained CLIP model to extract semantically rich image embeddings without the need for training a custom vision encoder.
    \item Utilize the generative capabilities of GPT-2 to produce fluent and coherent natural language captions.
    \item Bridge the gap between visual and textual modalities by introducing a projection layer that maps image embeddings into GPT-2's input space.
    \item Enable flexible and data-efficient image captioning, where the visual semantics guide the generation through prefix-based conditioning.
    \item Evaluate the quality of the generated captions using standard metrics such as BLEU and CIDEr, in order to quantitatively assess the model's accuracy and relevance.
\end{itemize}
\pagebreak
\section{Background and Related Work}

\subsection{CLIP(Contrastive Language-Image Pretraining)}
CLIP from OpenAI is a visual-language model. Instead of relying on 
task-specific supervised learning, CLIP is trained on a dataset of 400
 million image-text pairs collected from the internet using a contrastive 
 loss function. CLIP consists of two separate encoders: a visual encoder 
 (ResNet) for images, and a text encoder (Transformer) for captions. 
 Its ability to generate rich, semantically meaningful image 
 embeddings makes CLIP a powerful foundation for our systems and 
 an ideal visual component in our CLIP+GPT-2 image captioning pipeline. 
 For example, according to Mokady et al.\cite{Mokady2021}, it mentioned that the visual encoding capability of 
 CLIP can be used to embed and project the generated images into the input 
 space of GPT-2 to generate prefixes, which helps the final caption generation of GPT-2. Inspired by this article, we decided to study the CLIP architecture and implement related deployments.

 \subsection{CLIP(Contrastive Language-Image Pretraining)}
 GPT-2 is a large-scale language model based on the Transformer 
 decoder architecture proposed by Radford et al. \cite{Radford}. 
 According to the paper, Transformer completely replaces the traditional RNN or 
 CNN structure with self-attention, which is more efficient and accurate when processing 
 long sequence dependencies. Therefore, we consider implementing the transformer 
 structure as our decoder of the whole pipeline. Unlike the traditional 
 Transformer, which contains both encoder and decoder components, 
 GPT-2 uses only a decoder. This design enables the model to predict the next token based solely on previously generated tokens, 
 making the generated text semantically relevant and well suited for text generation tasks such as image captioning.

\begin{equation}
L = -\sum_{i \in I} \log \frac{\exp(\text{sim}(z_i, z_j)/\tau)}{\sum_{k \neq i}\exp(\text{sim}(z_i, z_k)/\tau)}
\end{equation}
where $\text{sim}(z_i, z_j)$ is the cosine similarity between embeddings, and $\tau$ is the temperature hyperparameter.

\subsection{Comparison to Classical Methods}
Prior to contrastive learning, traditional unsupervised feature learning methods included:
\begin{itemize}
  \item \textbf{Principal Component Analysis (PCA):} Projects data onto lower-dimensional eigenvectors
with maximal variance.
  \item \textbf{Autoencoders:} Learn representations by reconstructing input data through an encoder-decoder
framework.
\end{itemize}
We compare these classical approaches with SimCLR to understand its advantages and limitations.

\section{Methodology}

@@

\textbf{Augmentation Pipeline:} SimCLR relies heavily on augmentations, which include:
\begin{itemize}
  \item Random crop and resize
  \item Color jittering
  \item Gaussian blur
  \item Horizontal flipping
\end{itemize}

@@

    \bibitem{li2020oscar}
    Li, X., Yin, X., Li, C., Zhang, P., Hu, X., Zhang, L., Wang, L., Hu, H., Dong, L., Wei, F., Choi, Y., \& Gao, J. (2020). ``Oscar: Object-Semantics Aligned Pre-training for Vision-Language Tasks.'' ECCV 2020.

>>>>

Motivated by this progress, we focus our project on replicating a CLIP+GPT-2-based image captioning system, with inspriation from Nukrai et al.\cite{Nukrai2022}. Through this project, we seek to explore and deepen our understanding of 
machine learning, as well as its broader applications in fields such as computer vision and natural language generation.
\begin{itemize}
    \item Leverage the pretrained CLIP model to extract semantically rich image embeddings without the need for training a custom vision encoder.
    \item Utilize the generative capabilities of GPT-2 to produce fluent and coherent natural language captions.
    \item Bridge the gap between visual and textual modalities by introducing a projection layer that maps image embeddings into GPT-2's input space.
    \item Enable flexible and data-efficient image captioning, where the visual semantics guide the generation through prefix-based conditioning.
    \item Evaluate the quality of the generated captions using standard metrics such as BLEU and CIDEr, in order to quantitatively assess the model's accuracy and relevance.
\end{itemize}
\pagebreak
\section{Background and Related Work}

    \subsection{Contrastive Language-Image Pretraining(CLIP)}
    CLIP from OpenAI is a visual-language model. Instead of relying on 
    task-specific supervised learning, CLIP is trained on a dataset of 400
    million image-text pairs collected from the internet using a contrastive 
    loss function. CLIP consists of two separate encoders: a visual encoder 
    (ResNet) for images, and a text encoder (Transformer) for captions. 
    Its ability to generate rich, semantically meaningful image 
    embeddings makes CLIP a powerful foundation for our systems and 
    an ideal visual component in our CLIP+GPT-2 image captioning pipeline. 
    For example, according to Mokady et al.\cite{Mokady2021}, it mentioned that the visual encoding capability of 
    CLIP can be used to embed and project the generated images into the input 
    space of GPT-2 to generate prefixes, which helps the final caption generation of GPT-2. Inspired by this article, we decided to study the CLIP architecture and implement related deployments.

    \subsection{Generative Pre-trained Transformer 2 (GPT-2)}
    GPT-2 is a large-scale language model based on the Transformer 
    decoder architecture proposed by Radford et al.\cite{Radford}. 
    According to the paper, Transformer completely replaces the traditional RNN or 
    CNN structure with self-attention, which is more efficient and accurate when processing 
    long sequence dependencies. Therefore, we consider implementing the transformer 
    structure as our decoder of the whole pipeline. Unlike the traditional 
    Transformer, which contains both encoder and decoder components, 
    GPT-2 uses only a decoder. This design enables the model to predict the next token based solely on previously generated tokens, 
    making the generated text semantically relevant and well suited for text generation tasks such as image captioning.

    \subsection{Multi-Head Attention Mechanism}
    Inspired by the \textit{``Attention Is All You Need''} paper 
    \cite{vaswani2017attention}, we employ a multi-head attention 
    mechanism as a cross-modal connector between the CLIP image embedding
    and the GPT-2 language model. Instead of using a linear layer to directly 
    project the CLIP features into the GPT-2 embedding space, we employ an 
    attention mechanism that enables the model to selectively focus on 
    different aspects of the visual features when generating each token 
    in the caption, making the generated tokens more consistent with the 
    image information. Multi-head Attention network allows the model to 
    learn how different regions in the image affect language production.

We compare these classical approaches with SimCLR to understand its advantages and limitations.
\pagebreak
\section{Methodology}

@@

\textbf{Augmentation Pipeline:} SimCLR relies heavily on augmentations, which include:
\begin{itemize}
  \item Random crop and resize
  \item Color jittering
  \item Gaussian blur
  \item Horizontal flipping
\end{itemize}
\begin{equation}
    L = -\sum_{i \in I} \log \frac{\exp(\text{sim}(z_i, z_j)/\tau)}{\sum_{k \neq i}\exp(\text{sim}(z_i, z_k)/\tau)}
    \end{equation}
    where $\text{sim}(z_i, z_j)$ is the cosine similarity between embeddings, and $\tau$ is the temperature hyperparameter.

@@

    \bibitem{vaswani2017attention}
    Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N., Kaiser, Ł., \& Polosukhin, I. (2017). ``Attention is all you need.'' 
    Advances in Neural Information Processing Systems, 30.
